---
title: "spotifyr_analysis"
author: "Roberto Olayo"
date: "2/12/2021"
output: html_document
---

```{r}
library(tidyverse)
library(spotifyr)
library(knitr)
library(plotly)
```


# Authentication
```{r}
spot_credentials <- read_tsv('SPOTIFY_CREDENTIALS.txt')
SC_ID <- spot_credentials %>% filter(DATA_TYPE == 'SPOTIFY_CLIENT_ID') %>% select(VALUE)
SC_SE <- spot_credentials %>% filter(DATA_TYPE == 'SPOTIFY_CLIENT_SECRET') %>% select(VALUE)
```

```{r}
Sys.setenv(SPOTIFY_CLIENT_ID = SC_ID)
Sys.setenv(SPOTIFY_CLIENT_SECRET = SC_SE)
access_token <- get_spotify_access_token()
```

# What do o?
First I will gather all of our playlists
```{r}
user_info <- read_tsv('user_credentials.txt')

roberto_id <- user_info %>% 
  filter(USER=='Roberto') %>% 
  select(SPOTIFY_ID)

carina_id <- user_info %>% 
  filter(USER=='Carina') %>% 
  select(SPOTIFY_ID)

roberto_plays <- get_user_playlists(roberto_id, limit = 50)
carina_plays <- get_user_playlists(carina_id, limit = 50)
```
Now this info includes all of the playlists that we listen to, whether we where the ones that created them or not. Unfortunately, I cannot access our *Liked Songs*. So, to get a sense of our music tates, I will limit the analysis to the songs that are in the playlists that we created.

```{r}
roberto_id.str <- roberto_id %>% as.character()
carina_id.str <- carina_id %>% as.character()

roberto_plays <- roberto_plays %>% 
                  filter(owner.id == roberto_id.str) %>% 
                  select(id)
carina_plays <- carina_plays %>% 
                  filter(owner.id == carina_id.str) %>% 
                  select(id)
```

Now, I'm going to collect the songs that are within these playlists. I'll remove any duplicates in case there is a song added to more that one playlist (I know I sometimes do that hehe).

```{r}
fetch_our_songs <- function(playlist_ids){
  song_list <- lapply(playlist_ids$id , get_playlist_tracks)
  song.df <- bind_rows(song_list)
  song.df_unq <- song.df %>% distinct(track.id, .keep_all=TRUE)
  
  return(song.df_unq)
}

roberto_songs <- fetch_our_songs(roberto_plays)
carina_songs <- fetch_our_songs(carina_plays)
```

Now I will also include info about these songs, meaning their features

```{r}
fetch_songs_features <- function(song_ids){
  feature_list <- lapply(song_ids , 
                         function(x){get_track_audio_features(x)})
  song.df <- bind_rows(feature_list)
  
  song.df_re <- song.df %>% rename(track.id = id)
  
  return(song.df_re)
}

roberto_song_features <- fetch_songs_features(roberto_songs$track.id)
carina_song_features <- fetch_songs_features(carina_songs$track.id)
```

Joining the info 

```{r}
roberto_df <- roberto_songs %>% 
              left_join(roberto_song_features, by='track.id') %>% 
              add_column(User = 'Roberto')

carina_df <- carina_songs %>% 
             left_join(carina_song_features, by='track.id') %>% 
             add_column(User = 'Carina')

song_information <- bind_rows(list(roberto_df, carina_df))
```


Writing data to file since the last step took some time

```{r}
song_columns <- unlist(map(song_information, class))
selected_columns <- names(song_columns)[song_columns != "list"]

song_information_noList <- song_information %>% select(all_of(selected_columns))
write_tsv(song_information_noList, "roberto_carina_song_info.tsv")
```






