---
title: "download_data"
author: "Roberto Olayo"
date: "2/12/2021"
output: html_document
---

```{r}
library(tidyverse)
library(spotifyr)
library(knitr)
```


# Authentication
```{r}
spot_credentials <- read_tsv('SPOTIFY_CREDENTIALS.txt')
SC_ID <- spot_credentials %>% filter(DATA_TYPE == 'SPOTIFY_CLIENT_ID') %>% select(VALUE)
SC_SE <- spot_credentials %>% filter(DATA_TYPE == 'SPOTIFY_CLIENT_SECRET') %>% select(VALUE)
```

```{r}
Sys.setenv(SPOTIFY_CLIENT_ID = SC_ID)
Sys.setenv(SPOTIFY_CLIENT_SECRET = SC_SE)
access_token <- get_spotify_access_token()
```


# What do o?
First I will gather all of our playlists
```{r}
user_info <- read_tsv('user_credentials.txt')


playlist_gather <- function(user.id){
  
  user.str <- user.id %>% as.character()
  
  print(user.str)
  user.playlist.all <- get_user_playlists(user.id, limit = 50)
  
  user.playlist.filt <- user.playlist.all %>% 
    filter(owner.id == user.str) %>% 
    mutate(spotify.id=user.id) %>% 
    select(id, spotify.id)
  
  return(user.playlist.filt)
}

playlist_list <- lapply(user_info$SPOTIFY_ID, playlist_gather)

user_playlists <- bind_rows(playlist_list) %>% 
  rename(playlist.id = id) 

```


## Some manual additions  
For user "12126691763", there are a couple of playlists that are collected since they are collaborative. I add them manually here.
```{r}
additional.playlists <- data.frame(playlist.id = c("64ke8tO8xjQptgJ2HIGH1q", 
                                                   "4HUCJPIjYBrMqyd5ohEKjA",
                                                   "3DFrt1FvEhWt3ZvyhyD9Z5"),
                                   spotify.id = "12126691763")

user_playlists <- bind_rows(user_playlists, additional.playlists)
user_playlists
```


Now this info includes all of the playlists that we listen to, whether we where the ones that created them or not. Unfortunately, I cannot access our *Liked Songs*. So, to get a sense of our music tates, I will limit the analysis to the songs that are in the playlists that we created.

Now, I'm going to collect the songs that are within these playlists. I'll remove any duplicates in case there is a song added to more that one playlist (I know I sometimes do that hehe).

```{r}
fetch_our_songs <- function(playlist_ids){
  song_list <- lapply(playlist_ids$playlist.id , get_playlist_tracks)
  song.df <- bind_rows(song_list)
  song.df_unq <- song.df %>% distinct(track.id, .keep_all=TRUE)
  
  return(song.df_unq)
}

gather_songs <- function(spot.id, df=user_playlists){
  print(spot.id)
  user_playlists.spec <- df %>% 
    filter(spotify.id == spot.id) %>% 
    select(playlist.id)

  songs_df <- fetch_our_songs(user_playlists.spec)
  
  songs_df <- songs_df %>% 
    mutate(spotify.id = spot.id)
}

user_songs <- lapply(user_info$SPOTIFY_ID, gather_songs)

user_songs.df <- bind_rows(user_songs)
user_songs.df
```


To gather song features, I will create a combined list with all of our songs. This will potentially reduce the number of songs that have to be downloaded and will avoid having to download the same data twice. 


```{r}
all_songs.ids <- user_songs.df %>% 
  distinct(track.id)
paste("There are a total of", length(all_songs.ids$track.id), "distinct songs between the users")
```

```{r}
fetch_songs_features <- function(song_ids){
  feature_list <- lapply(song_ids , 
                         function(x){get_track_audio_features(x)})
  song.df <- bind_rows(feature_list)
  
  song.df_re <- song.df %>% rename(track.id = id)
  
  return(song.df_re)
}

all_songs.features <- lapply(all_songs.ids$track.id, fetch_songs_features)
all_songs.features.df <- bind_rows(all_songs.features)

length(unique(all_songs.features.df$track.id)) == length(all_songs.ids$track.id)

head(all_songs.features.df)
```


Joining the info 

```{r}
song_information <- user_songs.df %>% 
  left_join(all_songs.features.df, by="track.id") %>% 
  mutate(User=sapply(user_songs.df$spotify.id, function(x){user_info[user_info$SPOTIFY_ID==x, "USER"]$USER}))

tail(song_information)
```

Writing data to file since the last step took some time

```{r}
song_columns <- unlist(map(song_information, class))
selected_columns <- names(song_columns)[song_columns != "list"]

song_information_noList <- song_information %>% select(all_of(selected_columns))
write_tsv(song_information_noList, "user_song_information.tsv")
```


```{r}
song_information %>% 
  count(User) %>% 
  arrange(n)
```







